---
title: "Project1_Data612"
author: "Samriti Malhotra, Vishal Arora"
date: "June 7, 2019"
output:
  html_document:
    code_folding: hide
    highlight: tango
    number_sections: yes
    smooth_scroll: yes
    theme: united
    toc: yes
    toc_collapsed: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Building a basic Recommender system.
This is a basic recommender system where we created a sample ratings grid where movie(s) have been rated by users.Few vales are NA and hence this recommender system using BaseLine Predictor(avg mean + rowBias + columnB ias ) will try to predict the ratings for the ones which have not been rated and at the end we will check how much improvment it brings to train and test data set .

The libraries used during this project are as following
*reshape2*
*kableExtra*

```{r install-library-list, eval=TRUE, include=FALSE, echo=TRUE}
library(reshape2)
library(kableExtra)
```


```{r}
ratings <- read.csv("C:\\CUNY_AUG27\\Data612-RecommenderSystem\\Project1\\ratings.csv")
kable(head(ratings,n=9)) %>%
  kable_styling(bootstrap_options = c("striped","condensed","responsive"),full_width   = F,position = "left",font_size = 12) %>%
  row_spec(0, background ="gray")

#converting the ratings data frame into user-movie matrix 


ratingDT <- acast(ratings, user~movie, value.var="rating")
ratingDT <- apply(ratingDT, 2,as.numeric) 
kable(ratingDT) %>%
  kable_styling(bootstrap_options = c("striped","condensed","responsive"),full_width   = F,position = "left",font_size = 12) %>%
  row_spec(0, background ="gray")



## 75% of the sample size
smp_size <- floor(0.75 * nrow(ratingDT))

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(ratingDT)), size = smp_size)

train <- ratingDT[train_ind, ]
test <- ratingDT[-train_ind, ]

train
test


#deducing the raw mean average on train data set.
train.avg <- mean(train, na.rm =TRUE)
train.avg

#function to calculate RMSE
RMSE = function(data, data.avg){ 
  sqrt(mean((data - data.avg)^2, na.rm =TRUE)) 
}

trainRMSE_beforeBias <-  RMSE(train,train.avg)
trainRMSE_beforeBias


#Calculating the baseline predictor (raw average + userBias + movieBias)
calcBias <- function(dataMatrix, dataAvg){
  
  userBias <- rowMeans(dataMatrix, na.rm=T) - dataAvg
  movieBias <- colMeans(dataMatrix, na.rm=T) - dataAvg
  
  outMatrix <- dataMatrix
  rowcount <-1
  for(item in 1:nrow(dataMatrix))
  {
    colcount <-1
    for(colItem in 1: ncol(dataMatrix))
    {
      
     outMatrix[rowcount,colcount] <- dataAvg + userBias[[rowcount]] +  movieBias[[colcount]]
      colcount <- colcount +1  
    }
    rowcount <- rowcount +1
  }
  return (outMatrix)
}

baselineTrain <- calcBias(train,train.avg)


#baselineTrain[baselineTrain<0] <- 0 
#baselineTrain[baselineTrain>5] <- 5 


# Training dataset RMSE
trainRMSE_AfterBias <-RMSE(baselineTrain,train)

```



```{r}

# Raw mean of the test dataset
test.avg <- mean(test, na.rm =TRUE)

testRMSE_beforeBias <-  RMSE(test,test.avg)
testRMSE_beforeBias

# Test dataset RMSE
baselineTest <- calcBias(test, test.avg)

# Training dataset RMSE
testRMSE_afterBias <- RMSE(baselineTest,test)

```


#%Improvement for Train data set

```{r}
t1 <- trainRMSE_beforeBias
tb1 <- trainRMSE_AfterBias

trainImprove_perc <- (1-(tb1/t1))*100
trainImprove_perc
```

#%Improvement for Train data set
```{r}
t1 <- testRMSE_beforeBias
tb1 <- testRMSE_afterBias

testImprove_perc <- (1-(tb1/t1))*100
testImprove_perc
```

